The Post-2007 Declining U.S. Birth Rates imply Significant Societal Shifts in the U.S.

### 2023-02-20 outputs/paper/paper.qmd
Revised sections, Introduction and Data
Graphs and Tables are moved to under the result section.

---
title: "The Post-2007 Declining U.S. Birth Rates imply Significant Societal Shifts in the U.S."
author: 
  - Sagith Kalaichelvam
  - Youngho Kim
  - Joseph Chung
thanks: "Code and data are available at: https://github.com/UtopianYoungChung/Paper2.git"
date: "`r Sys.time()`"
date-format: "D MMMM YYYY"
abstract: "This paper explores several factors of the steep decline in birth rates in the United States between 2007 and 2020. The decline in U.S. birth rates offers some significant insights into U.S. society. Changes in birth patterns among recent young adult women moving through the childbearing age group can be directly related to the decline in birth rate from a purely statistical standpoint. In this paper, we decompose the decline in birth rates into between - and within - demographic groups. We then argue that any other economic, policy or social factor that has changed since 2007 does not contribute much to the decline. We also provide evidence that the steep drop in birth rates could be attributed to recent young adults' shifting interests. These shifting priorities represent changes in young adults' preferences regarding having children and aspirations for life. Resonating with the original paper, we conclude with a brief discussion of the implications of the declining birth rate in the United States and its impact on society as an entirety."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(palmerpenguins)
library(knitr)
library(ggplot2)
```

@@

We need Figure 1, Figure 2 - A & B, Table 1, Figure 4, and Figure 5 to show the evidence for our claim.

You can and should cross-reference sections and sub-sections. For instance, @sec-data and @sec-first-point.

@@

Figure 1. place it here with the Figure 1 from the Original paper (@fig-bills).

```{r}
#| label: fig-bills
#| fig-cap: Trend in US Birth Rates
#| echo: false

Talk more about it.

@@

Talk way more about it.

@@

Figure 4. place it here with the Figure 5 from the Original paper
```{r}
#| label: fig-billssss
#| fig-cap: Children Ever Born by Mother's Age and Birth Cohort
#| echo: false
#| fig-width: 8
#| fig-height: 4

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()

@@

# Model

$$
Pr(\theta | y) = \frac{Pr(y | \theta) Pr(\theta)}{Pr(y)}
$$ {#eq-bayes}

@eq-bayes seems useful, eh?

Here's a dumb example of how to use some references: In paper we run our analysis in `R` [@citeR]. We also use the `tidyverse` which was written by @thereferencecanbewhatever If we were interested in baseball data then @citeLahman could be useful.

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

>>>>

---
title: "The Post-2007 Declining U.S. Birth Rates imply Significant Societal Shifts in the U.S."
author: 
  - Sagith Kalaichelvam
  - Youngho Kim
  - Joseph Chung
thanks: "Code and data are available at: https://github.com/UtopianYoungChung/Paper2.git"
date: "`r Sys.time()`"
date-format: "D MMMM YYYY"
abstract: "This paper explores several potential explanatory factors for the steep decline in birth rates in the United States between 2007 and 2020. The decline in U.S. birthrates offers some significant insights into U.S. society. Changes in birth patterns among recent young adult women moving through the childbearing age group can be directly related to the decline in birth rate from a purely statistical standpoint. We provide evidence that the steep drop in birth rates could be attributed to recent young adults' shifting interests. These shifting priorities represent changes in young adults' preferences regarding having children and aspirations for life. Resonating with the original paper, we conclude with a brief discussion of the implications of the declining birth rate in the United States and its impact on society as an entirety."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(palmerpenguins)
library(knitr)
library(ggplot2)
library(haven)
library(dplyr)
options(dplyr.summarise.inform = FALSE)
```

@@

To measure birth rates, the National Center for Health Statistics (NCHS) aggregates data on birth rates for women by age, race, education level, and other factors; the aggregated data for a particular year is the total number of births per 1000 women. Each year NCHS measures birth rates nationwide and analyzes the data to make insights into societal phenomena. For instance, in 1997, NCHS published a paper on how a mother's education level influences the birth rate. In the paper, NCHS asserted that there is a negative correlation between a mother's education level and birth rate. Their study found a direct relationship between years of education and birth rates, with the highest birth rates among women with the lowest educational attainment (cite).

Figure 1 depicts U.S. birth rate trends over the last 40 years. In the figure, we can see that the trend is fluctuating relatively stable. However, during the post-2007 recession, it takes a steep decline with no signs of coming back up.

In this paper, we decompose the decline in the birth rate between - and within - demographic groups. We then argue that any other economic, policy or social factor that has changed since 2007 does not contribute much to the decline.

We next turn to a potential attributing factor for the declining U.S. birth rate after the 2007 recession. In our conclusion, we show strong evidence that the most recent group of young adults is significantly attributed to the post-2007 decline in the birth rate. This significance suggests an ongoing societal shift among the young women entering through childbearing age. Alarmingly, this societal phenomenon imposes some serious societal consequences and, therefore, should be considered at a high level. Most importantly, there are no signs of returning from the negative trend.

For instance, @sec-data and @sec-first-point.

@@

# Results {#sec-results}
Figure 1. Trends in US birth rates over the past 40 years show a steep decline in the birth rate during the post-2007 recession (@fig-1).

```{r}
#| label: fig-1
#| fig-cap: Trends in U.S. Birth rates of 1000 women between age 14 to 44 from 1980 to 2020
#| echo: false
fig1_data <- read.csv(here::here("inputs/data/fig_1.csv"))
fig1_data %>%
  ggplot(aes(x = year,
             y = brate_all)) +
  geom_line() +
  scale_y_continuous(limits = c(50, 80), breaks = seq(50, 80, by = 5)) +
  labs(
    title = "Trend in US Birth Rates",
    x = "Years",
    y = "Births per 1,000 women age 15-44",
    caption = "Source: Birth Rates collected from CDC Vital Statistics Births Reports for 2015, 2019 and 2020."
  )  +
  theme_light()
```

@@

@@

Figure 4.

```{r}
#| label: fig-4
#| fig-cap: cohorts
#| echo: false
nch_cohot_data <- read_dta(here::here("inputs/data/nchs/nchs_cohort_analysis.dta"))
agecomp_data <- read_dta(here::here("inputs/data/pop/agecomp-seer.dta"))

pop_by_cohort<- agecomp_data %>%
  select(1:32) %>%
  group_by(year) %>%
  select(-c(stname)) %>%
  summarise(across(everything(),sum)) %>%
  pivot_longer(-c(year), names_to = "mage", values_to = "pop") %>%
  mutate(mage = as.numeric(str_replace(mage, "fem", ""))) %>%
  mutate(birthyear = year - mage) %>%
  mutate(cohort2 = case_when(birthyear >= 1968 & birthyear <=1972 ~ 1,
                            birthyear >= 1973 & birthyear <=1977 ~ 2,
                            birthyear >= 1978 & birthyear <=1982 ~ 3,
                            birthyear >= 1983 & birthyear <=1987 ~ 4,
                            birthyear >= 1988 & birthyear <=1992 ~ 5,
                            birthyear >= 1993 & birthyear <=1997 ~ 6)) %>% 
  group_by(cohort2,mage) %>%
  summarise(pop = sum(pop))

fig5_data <- right_join(nch_cohot_data, pop_by_cohort, by=c("mage", "cohort2"))
fig5_data <- fig5_data %>%
  mutate(birthrate=numbirth/pop*1000,
         age_2024_year = case_when(cohort2 == 1 ~ 1992,
                                   cohort2 == 2 ~ 1997,
                                   cohort2 == 3 ~ 2002,
                                   cohort2 == 4 ~ 2007,
                                   cohort2 == 5 ~ 2012,
                                   cohort2 == 6 ~ 2017))

fig5_data <- fig5_data %>% 
  group_by(age_2024_year) %>% 
  mutate(cum_birthrate = cumsum(birthrate)/1000, 
         cohort = case_when(cohort2 == 1 ~ "1968-1972",
                                  cohort2 == 2 ~ "1973-1977",
                                  cohort2 == 3 ~ "1978-1982",
                                  cohort2 == 4 ~ "1983-1987",
                                  cohort2 == 5 ~ "1988-1992",
                                  cohort2 == 6 ~ "1993-1997"))

fig5_data <-  na.omit(fig5_data)
fig5_data %>%
  ggplot(aes(x = mage,
             y = cum_birthrate,
             color = cohort)) +
  geom_line()  +
  labs(
    title = "Children Ever Born by Mother’s Age and Birth Cohort",
    x = "ages",
    y = "cumulative birthrate",
    caption = "Source: Authors’ calculations based on NCHS Vital Statistics birth microdata and CDC SEER population
data. See the Data Appendix for specific details.",
    color = "Birth cohorts"
  ) +
  scale_x_continuous(limits=c(15, 44), breaks=seq(15, 44, by = 2)) +
  theme_light()

@@
